---
title: "CHEP 898 Supervised Learning"
author: "Harsimran Ghai"
date: "2026-03-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading the Library
```{r setup}
library(tidyverse)
library(tidymodels)
library(sjPlot)
library(C50)
library(ranger)
library(finalfit)
library(knitr)
library(gtsummary)
library(mlbench)
library(vip)
library(rsample)
library(rpart.plot)
library(tune)
library(recipes)
library(yardstick)
library(parsnip)
library(glmnet)
library(themis)
library(microbenchmark)
```

# Loading the Data
```{r}
data<-read_csv("Canpath_imputed.csv")
```

# Simplifying the dataset
```{r}
Small_data <- select(data, DIS_CROHN_EVER, HS_GEN_HEALTH, SDC_AGE_CALC, SDC_EDU_LEVEL, PM_BMI_SR, WRK_FULL_TIME, SMK_CIG_EVER, SDC_INCOME, PA_TOTAL_SHORT, HS_ROUTINE_VISIT_EVER, PSE_ADULT_WRK_DURATION, DIS_RESP_SLEEP_APNEA_EVER, SDC_EDU_LEVEL_AGE, SDC_GENDER)
```

# Preparing the variables
```{r}
table(Small_data$DIS_CROHN_EVER)

Small_data <- Small_data %>%
	mutate(Crohn = case_when(
		DIS_CROHN_EVER == 0 ~ "No",
		DIS_CROHN_EVER == 1 ~ "Yes",
		DIS_CROHN_EVER == 2 ~ "No")) %>%
		mutate(Crohn = as.factor(Crohn))

table(Small_data$DIS_CROHN_EVER, Small_data$Crohn)

Small_data$DIS_CROHN_EVER <- NULL
```

```{r}
table(Small_data$HS_GEN_HEALTH)

# Merging the level 1 and level 2 to make the level close in sample size

Small_data <- Small_data %>%
	mutate(gen_health = case_when(
		HS_GEN_HEALTH == 1 ~ 1,
		HS_GEN_HEALTH == 2 ~ 1,
		HS_GEN_HEALTH == 3 ~ 2,
		HS_GEN_HEALTH == 4 ~ 3,	
		HS_GEN_HEALTH == 5 ~ 4)) %>%
		mutate(gen_health = as.factor(gen_health))

table(Small_data$HS_GEN_HEALTH, Small_data$gen_health)

Small_data$HS_GEN_HEALTH <- NULL
```
# Exploratory data Analysis

## Changing variables into factors
```{r}
Small_data <- Small_data %>% mutate(across(c(SDC_EDU_LEVEL,WRK_FULL_TIME, SMK_CIG_EVER, SDC_INCOME, HS_ROUTINE_VISIT_EVER, DIS_RESP_SLEEP_APNEA_EVER,SDC_GENDER),as.factor))
```


## Descriptive of dataset
```{r}
summary(Small_data)
```
## Checking Duplicates
```{r}
duplicated(Small_data)
```
No Duplicated values were found in this dataset.

## Checking the missing values
```{r}
any(is.na(Small_data))
```
No missing values were found in the dataset as expected from imputed dataset.

# Splitting the dataset
```{r}
# Fixing the random numbers by setting the seed to reproduce the analysis
set.seed(10)

#### Cross Validation Split
cv_split <- initial_validation_split(Small_data, 
                            strata = Crohn, 
                            prop = c(0.70, 0.20))

# Create data frames for the two sets:
train_data <- training(cv_split)
table(train_data$gen_health)

test_data  <- testing(cv_split)
table(test_data$gen_health)
```

# Model Building

## Setting the core

```{r}
cores <- parallel::detectCores()
cores
```

```{r}
rf_model <- rand_forest(mtry = tune(), min_n = tune(), trees = 100) %>% 
              set_engine("ranger", num.threads = cores) %>% 
              set_mode("classification")
```

# Creating a Recipe

```{r}
Health_recipe <- 
  recipe(Crohn ~ ., data = train_data) %>% 
  step_zv(all_predictors()) 
```


# Creating a Workflow

```{r}
Health_workflow <- 
        workflow() %>% 
        add_model(rf_model) %>% 
        add_recipe(Health_recipe)

Health_workflow
```

# Fitting the Model

```{r}
set.seed(100)

folds <- vfold_cv(train_data, v = 10) 

rf_grid <- grid_regular(
              mtry(range = c(2, 13)),
              min_n(range = c(5, 50)),
              levels = 5  
            )

rf_grid


Health_fit <- tune_grid(
                Health_workflow,
                resamples = folds,
                grid = rf_grid, 
                control = control_resamples(save_pred = TRUE, 
                                                  verbose = FALSE))

Health_fit
```

# Plotting of the results. 

```{r}
Health_fit %>%
  collect_metrics() %>%
  filter(.metric == "accuracy") %>%
  select(mean, min_n, mtry) %>%
  pivot_longer(min_n:mtry,
    values_to = "value",
    names_to = "parameter"
  ) %>%
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "Accuracy")
```

```{r}
Health_fit %>%
  collect_metrics() %>%
  filter(.metric == "accuracy") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point() +
  labs(y = "Accuracy")
```
### Collecting the results

```{r}
rf_best <- 
  Health_fit %>% 
  select_best(metric = "accuracy")

rf_best

rf_auc_fit <- 
  Health_fit %>% 
  collect_predictions(parameters = rf_best) 
```

# Metrics

## Confusion Matrix

```{r}
conf_mat(rf_auc_fit, truth = Crohn,
         estimate = .pred_class)
```

# Accuracy

```{r}
accuracy(rf_auc_fit, truth = Crohn,
         estimate = .pred_class)
```

```{r}
sens(rf_auc_fit, truth = Crohn,
         estimate = .pred_class)
```

```{r}
spec(rf_auc_fit, truth = Crohn,
         estimate = .pred_class)
```
```{r}
f_meas(rf_auc_fit, truth = Crohn,
         estimate = .pred_class)
```
# Final Model 

```{r}
final_model <- finalize_model(
                  rf_model,
                  rf_best
                )

final_model
```
### Variable Importance

```{r}
tree_prep <- prep(Health_recipe)

final_model %>%
  set_engine("ranger", importance = "permutation") %>%
  fit(Crohn ~ .,
    data = juice(tree_prep)) %>%
  vip(geom = "point")
```

```{r}
final_rf_workflow <- workflow() %>%
                      add_recipe(Health_recipe) %>%
                      add_model(final_model)

final_results <- final_rf_workflow %>%
                    last_fit(cv_split)

final_results %>%
  collect_metrics()
```